Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material

### 2020-05-20 paper/response.tex

# Reviewer 2

First, the authors justify their approach of using backward modeling by the
    ability to estimate common noise sources. However, this would also be possible
    using multivariate forward modeling. For that reason, I'm wondering how
    multivariate forward modeling would fare under these circumstances. I would find
    it highly informative to include a (similarly cross-validated) multivariate GLM
    to the analyses and results.

We thank Reviewer 2 for this remark. The multivariate GLM / MANOVA is indeed
a relevant analysis. We now indicate in supplementary analyses:

Following the recommendations of one of our reviewers, we also implemented a
multivariate variant of the forward model, i.e. a MANOVA, using the statsmodels
implementation \citep{seabold2010statsmodels}. MANOVA is primarily used as a
inferential statistics, and does not trivially convert to a predicting method.
Consequently, we did not find a way to compare MANOVA against B2B with the
$\Delta R$ evaluation. However, the effects of MANOVA comes with its own effect
size statistics: the Wilk's Lambda. For each searchlight, we thus use the $F$-values of the
Wilk's Lambda statistics as proxy for $\hat S$ and feeds it to a second-level
Wilcoxon tests across subjects, like we did for the other models.

The results show that all factors, including the Dummy variable, are
systematically above chance level in all recorded brain regions (Supplementary Fig.~\ref{fig:fmri_supp}). This result
can be explained by the large dimensionality of $Y$. Indeed, limiting the
searchlight to a 1mm radius did not lead to these spurious effects but provided
results similar to the Forward model.

Nonetheless, MANOVA does appear to capture some plausible effects. Indeed, the
$F$-values obtained for both Word Length and Word Frequency were weakly but
significantly higher than those obtained with the Dummy variable in the
occipital and temporal brain areas~\ref{fig:fmri_supp}.

@@

It is possible to report E_hat (now relabelled S_hat). We now added a Supplementary Figure
to that effect. Note, however, that E_hat statistics are not in the same unit as the forward H_hat
coefficients (a.k.a betas values in GLM). Consequently, it is not straightforward to
compare these two metrics directly. As a proxy, we thus report their respective
second-level p-values across subjects. This figure shows show that B2B S_hat
coefficients compare favorably to the Forward H_hat coefficient and the MANOVA coefficients.

- The authors describe their method as unbiased but then introduce two
    regularization parameters, which will bias the estimated coefficients towards
    zero. While the statement "B2B leads to unbiased (i.e. zeros-centered) scalar
    coefficients for non-causal features" is correct for the mean of those
    coefficients, the variance will also be biased. I think these are important
    criteria to consider.

We agree with this remark. We now clarify the meaning and implication of the bias.

Note that $\hat S$ is unbiased, in the sense that it is centered around zero when there is no
effect, only if the second regression $H$ is not regularized. Second-level
statistics testing whether $\hat S$ is superior to 0 are thus only valid if $H$
is not regularized.

### 2020-05-20 paper/response.tex

Reviewers' comments:

Reviewer #1:
    There is one major limitation for the broader application of this approach and
    that's demonstrating its effectiveness in fMRI data. The majority of
    neuroimaging studies are done using fMRI (and I do not say this to discount the
    other valuable tools of EEG, MEG, NIRS, etc.). To get other neuroscientists to
    grab onto this approach, I recommend the authors perform another analysis using
    fMRI data. My suggestion for the authors would be to use data from the Human
    Connectome Project (http://www.humanconnectomeproject.org/). The outcome
    variable used with this approach is not important as so much the demonstration
    that their technique is effective for fMRI data. Without including fMRI data,
    the approach comes off as useful but speculative for fMRI data, which
    drastically limits its impact potential.

We thank Reviewer #1 for this remark. We now apply the same analyses to
fMRI recordings of 100 subjects performing a reading experiment similar to the
MEG study described in original manuscript. We kept the exact same methods and parameters, and
used the fmriprep package to perform the preprocessing.

Overall, the results consistently show that B2B compares favorably to other methods.
Furthermore, these novel analyses complements the temporal results obtained with
MEG. Specifically, the effects of Word Length are predominantly observed over the early visual
cortices, and the Word Frequency are predominantly observed over the fronto-temporal areas.

# Reviewer 2:
    First, the authors justify their approach of using backward modeling by the
    ability to estimate common noise sources. However, this would also be possible
    using multivariate forward modeling. For that reason, I'm wondering how
    multivariate forward modeling would fare under these circumstances. I would find
    it highly informative to include a (similarly cross-validated) multivariate GLM
    to the analyses and results.

We thank Reviewer 2 for this remark. The multivariate GLM / MANOVA is indeed
a relevant analysis. We now indicate in supplementary materials the following analyses:

"
Following the recommendations of one of our reviewers, we implemented a
multivariate variant of the forward model, i.e. a MANOVA, using the statsmodels
implementation \citep{seabold2010statsmodels}. MANOVA is primarily used as a
inferential statistics, and does not trivially convert to a predicting method.
Consequently, we did not find a way to compare MANOVA against B2B with the
$\Delta R$ evaluation. However, the effects of MANOVA are generally summarized
with the Wilk's Lambda stastistics or its transformation into an $F$-value.
For each searchlight, we thus use the $F$-values of the
Wilk's Lambda statistics as proxy for $\hat S$ and feeds it to a second-level
Wilcoxon signed-rank tests across subjects, like we did for the other models.

The results show that all factors, including the Dummy variable, are
systematically above chance level in all recorded brain regions (Fig.~\ref{fig:fmri_supp}). This result
can be explained by the large dimensionality of $Y$. Indeed, limiting the
searchlight to a 1mm radius did not lead to these spurious effects but provided
results similar to the Forward model.

Nonetheless, MANOVA does appear to capture some plausible effects. Indeed, the
$F$-values obtained for both Word Length and Word Frequency were weakly but
significantly higher than those obtained with the Dummy variable in the
occipital and temporal brain areas (Fig.~\ref{fig:fmri_supp}). This results
suggests that the effect size of the MANOVA can be biased and, thus, is not
valid for second-level statistics.

Overall, this suggests that MANOVA (1) can lead to positively biased estimates of
$\hat S$ (2) appears weaker than B2B in terms of second-level analysis across
subjects (3) misses the effect of Word Function detected with the Forward model,
and (4) does not trivially translate into a prediction tool. Together these
elements thus suggest that MANOVA is less suitable to the present objective than
B2B.
"

@@

We agree with Reviewer 2. As indicated to Reviewer 1, we now successfully apply
the same analyses to the fMRI recordings of 100 subjects reading word sequences.

Other comments:

- Rather than using the feature importance delta_R, is it possible to just
    report the parameter estimate E_hat at each time point? What would be the
    downside of this?

It is possible to report $\hat E$ (now relabelled $\hat S$), however they are
interpretable only if the second regression $H$ is not regularized.
We now added a Supplementary Figure to that effect.

Note, however, that E_hat statistics are not in the same unit as the forward H_hat
coefficients (a.k.a betas values in GLM). Consequently, it is not straightforward to
compare these two metrics directly. As a proxy, we thus report their respective
second-level p-values across subjects. The associated figure shows that B2B's $\hat S$
coefficients compare favorably to the Forward H_hat coefficient and the MANOVA
coefficients.

@@

- The authors describe their method as unbiased but then introduce two
    regularization parameters, which will bias the estimated coefficients towards
    zero. While the statement "B2B leads to unbiased (i.e. zeros-centered) scalar
    coefficients for non-causal features" is correct for the mean of those
    coefficients, the variance will also be biased. I think these are important
    criteria to consider.

We agree with this remark. We now clarify the meaning and implication of the
bias in the problem statement:

"Note that $\hat S$ is unbiased, in the sense that it is centered around zero when there is no
effect, only if the second regression $H$ is not regularized. Second-level
statistics testing whether $\hat S$ is superior to 0 are thus only valid if $H$
is not regularized."

@@

- The results of backward regression are not shown in Figure 4, but it would be
    nice to see those results. Is there a possibility to add them to Figure 4 in a
    separate panel?

We now added the decoding performance to a supplementary figure. The latter show
that approximately the same cortical network can be used to decode Word Length
and Word Frequency. It also shows that the Dummy Variable can be decoded from
various brain areas, and from the visual cortex in particular.

@@

- I think the authors could better explain the dimensions of the data (dy
    referring to the number of measurement channels, and dx to the number of
    independent variables). Perhaps the authors could add this nomenclature to their
    graphical illustration? Likewise, for the regression of Y onto X, it is unclear
    if this is a mass-univariate regression or a multivariate regression (i.e.
    taking the error covariance between Ys into account or not). This might be
    obvious to them, but is not necessarily obvious to the reader.

We amended the text in multiple places to clarify this issue.

### 2020-05-20 paper/_check_can_be_removed_supplementary.tex
clean up

\newpage
\appendix
\section{Appendix}

\subsection{Proof of consistency theorem}
\label{appendix:theorem_proof}

Proof of the theorem in \ref{sec:theorem}:

@@

     Consider the B2B model from equation \ref{eq:model} $$Y = (XS + N)F$$ with
     $N$ centered and full rank noise.

     If $F$ and $X$ are full-rank on $Img(S)$, then, the solution of B2B, $\hat
     H$ minimizes

     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfies

@@

In the same way, for $\hat H$, we have
\begin{align*}
    \hat H = \arg \min_H \mathbb{S}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{S}[\| XH - (XS + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{S}[\| X(H - SF \hat G) \| ^2] + \mathbb{S}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{S}[\| X(H - SF \hat G) \| ^2]
 \end{align*}
 a positive quantity which reaches a minimum (zero) for
 \begin{align}
    \hat H = SF \hat G
    \label{eq:Hdoublenom}
\end{align}

@@

\newpage
\subsection{Modeling measurement noise}

Equation \ref{eq:model} does not explicitly contain a measurement noise term.
Yet, in most experimental cases, the problem is best described as:
\begin{equation}
  Y = (XS+N)F+M
  \label{eq:model_noisy_measure}
\end{equation}
  with $M\in \mathbb{R}^{m \times d_y}$.

This equation is actually equivalent to Equation \ref{eq:model} given our
hypotheses. Indeed, we can rewrite $M = MF^{-1}F$ over $Img(F)$, which leads to:
$$Y = (XS+N)F+M = (XS+N+MF^{-1})F = (XS+N')F$$

Consequently, assuming that $F$ is full rank on $Img(XS)$, B2B yields the same
solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.

\subsection{Feature importance}
\label{appendix:feature_importance}

For B2B, feature importance is assessed as follows:

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\;
    $R_{k} =
\text{corr}(X_{test} K H, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:b2b_fi}
\end{algorithm}

For the Forward Model, the feature importance is assessed as follows:

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.}

$H = \text{LinearRegression}(X_{train}, Y_{train})\;
R_{full} = \text{corr}(X_{test} K, Y_{test})$\;

    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $R_{k} = \text{corr}(X_{test}
K H, Y_{test})$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
\caption{Forward feature importance.} \label{algorithm:fwd_fi} \end{algorithm}

% \iffalse
% \begin{enumerate}
%   \item Fit $H$ given $X_{train}$ and $Y_{train}$
%   \item For each feature $i$:
%   \begin{itemize}
%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
%     \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train}$
%   \end{itemize}
%   \end{enumerate}
% \fi

For the CCA and PLS models, the feature importance is assessed as follows:

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x, d_z}$.} $H, G = \text{CCA}(X_{train},
Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $R_{k} = \text{corr}(X_{test} K H,
Y_{test} G)$\; $\Delta R_i = R_{full} - R_{k}$\; } \Return{$\Delta R$}
\caption{CCA and PLS feature importance.} \label{algorithm:cdp_fi}
\end{algorithm}

For the Backward Model, feature importance cannot be assessed because there is
no prediction.

\subsection{Recovering S}
\label{recovering}

In case of noise, B2B yields non binary $\hat S$. Three thresholding rules can
be used to binarize its values thus explicitly recover "causal" features.

First, given known signal-to-noise ratio, the threshold above which a feature
should considered to be "causal" can be derived analytically. Indeed, Equation
$\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are
bounded:  % we assume that $F$ has full rank over $Img(S)$.

$$0 \leq \frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq
\text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
where $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote
the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1
N_1}$.

The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the
trace of $\hat H$ divided by $k$, and can be computed as
\begin{equation}
\mu = \frac{Var(X)}{Var(X)+Var(N)}
\end{equation}

The decision threshold between "causal" and "non-causal" elements is thus a fraction
$\mu$, whose proportion arbitrarily depends on the necessity to favor type I and
type II errors. In practice, we cannot use this procedure for our MEG study,
because signal-to-noise ratio is unknown.

Second, $\text{diag}(\hat H)$ can be binarized with the Sonquist-Morgan
criterion \citep{sonquist_morgan}, a non-parametric clustering procedure
separating small and large values in a given set. This procedure maximizes the
ratio of inter-group variance while minimizing the intra-group variance, over
all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest
values. Let $m_0$ and $m_1$ be the average values of the two clusters, $p$ and
$d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan
criterion maximizes \citep{kass75}:
\begin{equation}
  \frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}
\end{equation}

This procedure assumes that there exists at least one causal and at least one
non-causal feature.
%
Third, second-order statistics across multiple datasets can be used to identify
the elements of $\text{diag}(\hat H)$ that are significantly different from 0.
This procedure is detailed in the method section of our MEG experiment.

Overall, these three procedures thus vary in their additional assumptions: i.e.
(1) a known signal-to-noise ratio, (2) the existence of both causal and
non-causal factors or (3) independent repetitions of the experiment.

\section{Additional Figures}

% \begin{figure}[t!]
%   \centering
%   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
%   \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects. Formally, Backward model cannot have a $\Delta R$ because it never unmixes the multiple $X$ features. In such cases, we thus simply report the decoding score.}
%   \label{fig:meg_supp}
% \end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/delta_auc.pdf}
  \caption{Synthetic experiments. Distribution (over conditions) of AUC (top) and Feature
  Importance $\Delta R$ (bottom) metrics between our method (y-axis) and the baselines
  (x-axis). Each dot is a distinct synthetic experiment. Dots below the diagonal indicates
  that B2B outperform the tested model. \label{fig:auc_plots}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{figures/duration}
  \caption{Wall-clock run-time for our method B2B and for the baselines. Each dot is a
  distinct synthetic experiment. B2B runs much faster than cross-decomposition baselines.
  \label{fig:duration}}
\end{figure}

\subsection{Supplementary comparison of models' coefficients}

Following the recommendations of one of our reviewers, we also implemented a
multivariate variant of the forward model, i.e. a MANOVA, using the statsmodels
implementation \citep{seabold2010statsmodels}. MANOVA is primarily used as a
inferential statistics, and does not trivially convert to a predicting method.
Consequently, we did not find a way to compare MANOVA against B2B with the
$\Delta R$ evaluation. However, the effects of MANOVA comes with its own effect
size statistics: the Wilk's Lambda. For each searchlight, we thus use the $F-values$ of the
Wilk's Lambda statistics as proxy for $\hat S$ and feeds it to a second-level
Wilcoxon signed-rank tests across subjects, like we did for the other models.

The results show that all factors, including the Dummy variable, are
systematically above chance level in all recorded brain regions \ref{fig:fmri_supp}. This result
can be explained by the large dimensionality of $Y$. Indeed, limiting the
searchlight to a 1mm radius did not lead to these spurious effects but provided
results similar to the Forward model.

Nonetheless, MANOVA does appear to capture some plausible effects. Indeed, the
$F-values$ obtained for both Word Length and Word Frequency were weakly but
significantly higher than those obtained with the Dummy variable in the
occipital and temporal brain areas \ref{fig:fmri_supp}.

Overall, this suggests that MANOVA (1) can lead to positively biased estimates of
$\hat S$ (2) appears weaker than B2B in terms of second-level analysis across
subjects (3) misses the effect of Word Function detected with the Forward model,
and (4) does not trivially translate into a prediction tool. Together these
elements thus suggest that MANOVA is less suitable to the present objective than
B2B.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.7\textwidth,
                     trim=0cm 0cm 0cm 0cm,
                     clip=True]{figures/fmri_controls.pdf}

    \label{fig:fmri_supp}
  \end{center}
  \caption{Top rows. Pearson $R$ correlation obtained for a Backward decoding model and
  a Forward encoding model trained with one factor at a time. These models can
  not take into account the factor covariance, and thus lead to spurious effects
  (e.g. visual cortex effect for the dummy variable). Bottom rows. Second-
  level p-values across subjects for a Forward, a B2B and a Manova trained with
  all factors. The B2B achieves better p-values, without leading to spurious effects for
  the Dummy variable. The Manova leads to biased estimates due to its inability
  to deal with overfitting. The last row shows where the Manova F-values of each factor differs
  significantly from the F-values of the Dummy variable.}
\end{figure}

\subsection{Robustness to increasing number of factors}

To test whether each of the methods robustly scales to an increasingly
large number of potential causes $X$, we enhanced the four ad-hoc features
(word length, word frequency, word function, dummy variable) with another
ten features. These additional features corresponds to the first dimensions
of word embedding as provided by Spacy \citep{spacy2}. The results shown
in Figure \ref{fig:embeddings}, show that the feature importance of ad-hoc
features as derived by B2B remain unchanged and are actually improved.

\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{figures/compare_embeddings.pdf}
  \caption{Comparison of $\Delta R$ when the models are tested on four
  variables (top) and when the models are tested on an these four variables
  as well as another 10 word-embedding features (bottom). These results
  illustrate that, unlike Regularized CCA, B2B remains robust even when
  the number of tested factors increases.
  \label{fig:embeddings}}
\end{figure}

>>>>

\documentclass{article}

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage[]{neurips_2019}
\newtheorem{theorem}{Theorem}

\title{Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material}

\begin{document}

\appendix

\maketitle

\section{Theorem - detailed proof}
\label{sec:theorem}

@@

     Consider the B2B model from Equation $$Y = (XS + N)F$$ $N$ centered and full rank noise.

     If $F$ and $X$ are full-rank on $Img(S)$, then, the solution of B2B, $\hat H$, will minimize

     $$\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$$ and satisfy

@@

Samely, for $\hat H$, we have
\begin{align*}
    \hat H = \arg \min_H \mathbb{S}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{S}[\| XH - (XS + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{S}[\| X(H - SF \hat G) \| ^2] + \mathbb{S}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{S}[\| X(H - SF \hat G) \| ^2]
 \end{align*}
 a positive quantity which reaches a minimum (zero) for
 \begin{align}
    \hat H = SF \hat G
    \label{eq:Hdoublenom}
\end{align}

@@

\end{document}
